Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-20 final_metrics_analysis.Rmd
more graph interpretations, fixing a few typos

---
title: "Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm"
author: "Elisabeth Nesmith, Eunice Kim, Emma Kornberg, Amrita Acharya, Dianne Caravela"
date: "3/8/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

@@

def plot_fair_metrics(fair_metrics_mitigated, model): 
  cols = ['statistical_parity_difference','equal_opportunity_difference','average_odds_difference','disparate_impact_ratio']
  obj_fairness = [[0,0,0,1]]

  # row for objectives    
  fair_metrics = pd.DataFrame(data=obj_fairness, index=['objective'], columns=cols)

  # row for baseline model
  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]

  # row for mitigated bias
  fair_metrics.loc[model] = fair_metrics_mitigated

  metrics_len = len(cols)

  fig, ax = plt.subplots(figsize=(20,4), ncols=metrics_len, nrows=1)

  plt.subplots_adjust(
      left    =  0.125, 
      bottom  =  0.1, 
      right   =  0.9, 
      top     =  0.9, 
      wspace  =  .5, 
      hspace  =  1.1
  )

  y_title_margin = 1.2

  plt.suptitle("Fairness metrics", y = 1.09, fontsize=20)
  sns.set(style="dark")

  cols = fair_metrics.columns.values
  obj = fair_metrics.loc['objective']
  size_rect = [0.2,0.2,0.2,0.4]
  rect = [-0.1,-0.1,-0.1,0.8]
  bottom = [-1,-1,-1,0]
  top = [1,1,1,2]
  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.25]]

  for i in range(0,metrics_len):
      plt.subplot(1, metrics_len, i+1)
      ax = sns.barplot(x=fair_metrics.index[1:len(fair_metrics)], y=fair_metrics.iloc[1:len(fair_metrics)][cols[i]])

      for j in range(0,len(fair_metrics)-1):
          a, val = ax.patches[j], fair_metrics.iloc[j+1][cols[i]]
          marg = -0.2 if val < 0 else 0.1
          ax.text(a.get_x()+a.get_width()/5, a.get_y()+a.get_height()+marg, round(val, 3), fontsize=15,color='black')

      plt.ylim(bottom[i], top[i])
      plt.setp(ax.patches, linewidth=0)
      ax.add_patch(patches.Rectangle((-5,rect[i]), 10, size_rect[i], alpha=0.3, facecolor="green", linewidth=1, linestyle='solid'))
      plt.axhline(obj[i], color='black', alpha=0.3)
      plt.title(cols[i])
      ax.set_ylabel('')    
      ax.set_xlabel('')
```

@@

X_new_index.groupby(["def_race"])["age"].median()
X_new_index.groupby(["def_race", "sex"]).size()
X_new_index.groupby(["def_race", "c_charge_degree"]).size()
X_new_index.groupby(["def_race"])["priors_count"].median()
#X_new_index.groupby(["def_race"])["juv_fel_count"].median()
#X_new_index.groupby(["def_race"])["juv_misd_count"].median()
# above two lines commented out since medians for everyone = 0

```

@@

This plot shows the distribution of two year recidivism outcomes. Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2700 did not.

@@

This plot shows the distribution of defendant races. The dataset contains about 3200 Black defendants and 2100 white defendants. Therefore, there are more Black people represented in the dataset than white people, showing the disproportionate policing and criminalization of Black communities.

```{python exploratory age by race plot}
# age distribution by race
#ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
#kdeline = ax.lines[0]
#mean_black = df_viz.groupby('race').age.median()[0]
#xs = kdeline.get_xdata()
#ys = kdeline.get_ydata()
#height_black = np.interp(mean_black, xs, ys)
#ax.vlines(mean_black, 0, height_black, color=purple, ls=':')

#mean_white = df_viz.groupby('race').age.median()[1]
#height_white = np.interp(mean_white, xs, ys)
#ax.vlines(mean_white, 0, height_white, color=green, ls=':')

#print(df_viz.groupby('race').age.median())

#plt.title('Distribution of Age by Race')
```

This plot shows the distribution of defendant ages by race. The purple curve shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years). For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.

@@

This plot shows the distribution of two year recidivism outcomes by race. When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. Again, this trend points to disproportionate rates of criminalization of Black people. This increased likelihood of recidivating may be attributed to the systemic racism in policing (e.g., predominantly Black neighborhoods tend to be overpoliced in comparison to predominantly white neighborhoods).

@@

This table shows the counts of recidivism by race.

@@

plt.title('Priors Count by Race')
plt.show()
```

@@

plt.title('Priors Count by Sex')
plt.show()
```

The two Priors Count bar charts show the differences in the distribution of priors by race and sex. If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism. In the former plot, one can see that Black defendants have a higher number of priors than do white defendants. This again likely reflects the over-policing of Black neighborhoods. In the latter plot, one can see that male defendants have a higher number of priors than do female defendants. This is worth noting because the COMPAS risk assessment algorithm has been found to express a discrepancy in prediction between sexes. When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes.

@@

$$\frac{TP}{TP + FP} $$ 
We can see that precision is very similar across groups, all approximately 0.6. This means that the model's predictions of a defendant, either Black or white, who recidivates are accurate about 60% of the time.

White people have a lower recall rate, meaning that there is a low true positive rate. Whereas for black people the true positive rate is much higher: there's a high percentage of recidivated guesses that were correct.

@@

@@

This metric returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 is considered fair, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both black and white defendants.

$$\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}$$

@@

$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$

```{python group fairnesss equal opportunity difference}
eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)

@@

## Reweighing

@@

The highest percentage in this confusion matrix is for the true negative rate, those who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model. This means that the re-weighing model marginally improved the baseline model accuracy. The true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model. This means that the re-weighing model slightly lowered the accuracy of the baseline model. The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively. The re-weighing model lowered the accuracy of the false negative rate by 1.14% and improved the accuracy of the false positive rate by 0.83%. We are most concerned about the false positivesthat mean that defendants who are predicted to recidivate and do not actually recidivate. While there is a slight accuracy improvement in the false positive rate, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.

@@

@@

```{python}
def get_attributes(data, selected_attr=None):
    unprivileged_groups = []
    privileged_groups = []
    if selected_attr == None:
        selected_attr = data.protected_attribute_names

    for attr in selected_attr:
            idx = data.protected_attribute_names.index(attr)
            privileged_groups.append({attr:data.privileged_protected_attributes[idx]}) 
            unprivileged_groups.append({attr:data.unprivileged_protected_attributes[idx]})

    return privileged_groups, unprivileged_groups
```

```{python post-processing equal odds needs fix} 
privileged_groups, unprivileged_groups = get_attributes(train_BLD, selected_attr= 'race')
ep = EqOddsPostprocessing(unprivileged_groups= unprivileged_groups, privileged_groups= privileged_groups, seed = 42)
#eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
EOPP = ep.fit(train_BLD, test_BLD)
y_pred_EO = ep.predict(test_BLD)
#y_proba_EO = ep.predict_(test_BLD)

acc_EO = accuracy_score(y_test, y_pred_EO)
```

>>>>

---
title: "Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm"
author: "Elisabeth Nesmith, Eunice Kim, Emma Kornberg, Amrita Acharya, Dianne Caravela"
date: "3/8/2022"
output: html_document
editor_options: 
  chunk_output_type: inline
---

@@

def plot_fair_metrics(fair_metrics_mitigated, model): 
  cols = ['statistical_parity_difference','equal_opportunity_difference','average_odds_difference','disparate_impact_ratio']
  obj_fairness = [[0,0,0,1]]
  # row for objectives    
  fair_metrics = pd.DataFrame(data=obj_fairness, index=['objective'], columns=cols)

  # row for baseline model
  fair_metrics.loc['Baseline Model'] = [stat_par_diff, eq_opp_diff, avg_odds_diff, disp_impact_ratio]
  # row for mitigated bias
  fair_metrics.loc[model] = fair_metrics_mitigated
  metrics_len = len(cols)
  fig, ax = plt.subplots(figsize=(20,4), ncols=metrics_len, nrows=1)
  plt.subplots_adjust(
      left    =  0.125, 
      bottom  =  0.1, 
      right   =  0.9, 
      top     =  0.9, 
      wspace  =  .5, 
      hspace  =  1.1
  )
  y_title_margin = 1.2
  plt.suptitle("Fairness metrics", y = 1.09, fontsize=20)
  sns.set(style="dark")
  cols = fair_metrics.columns.values
  obj = fair_metrics.loc['objective']
  size_rect = [0.2,0.2,0.2,0.4]
  rect = [-0.1,-0.1,-0.1,0.8]
  bottom = [-1,-1,-1,0]
  top = [1,1,1,2]
  bound = [[-0.1,0.1],[-0.1,0.1],[-0.1,0.1],[0.8,1.25]]
  for i in range(0,metrics_len):
      plt.subplot(1, metrics_len, i+1)
      ax = sns.barplot(x=fair_metrics.index[1:len(fair_metrics)], y=fair_metrics.iloc[1:len(fair_metrics)][cols[i]])

      for j in range(0,len(fair_metrics)-1):
          a, val = ax.patches[j], fair_metrics.iloc[j+1][cols[i]]
          marg = -0.2 if val < 0 else 0.1
          ax.text(a.get_x()+a.get_width()/5, a.get_y()+a.get_height()+marg, round(val, 3), fontsize=15,color='black')
      plt.ylim(bottom[i], top[i])
      plt.setp(ax.patches, linewidth=0)
      ax.add_patch(patches.Rectangle((-5,rect[i]), 10, size_rect[i], alpha=0.3, facecolor="green", linewidth=1, linestyle='solid'))
      plt.axhline(obj[i], color='black', alpha=0.3)
      plt.title(cols[i])
      ax.set_ylabel('')    
      ax.set_xlabel('')
```

@@

X_new_index.groupby(["def_race"])["age"].median()
X_new_index.groupby(["def_race", "sex"]).size()
X_new_index.groupby(["def_race", "c_charge_degree"]).size()
X_new_index.groupby(["def_race"])["priors_count"].median()
X_new_index.groupby(["def_race"])["juv_fel_count"].median()
X_new_index.groupby(["def_race"])["juv_misd_count"].median()
```

The typical Black defendant is 29 years old, male, and has committed two prior crimes. The typical white defendant is 35 years old, male, and has committed one prior crime. The typical defendant for both races does not have any juvenile convictions.

@@

Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2700 did not.

@@

The dataset we are using contains about 3200 Black defendants and 2100 white defendants. Therefore, there are more Black people represented in the dataset than white people, showing the disproportionate policing and criminalization of Black communities.

```{python exploratory age by race plot}
# age distribution by race - hacking solution for purple dotted line for now
ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
kdeline = ax.lines[0]
mean_black = df_viz.groupby('race').age.median()[0]
xs = kdeline.get_xdata()
ys = kdeline.get_ydata()
height_black = 0.027 # np.interp(mean_black, xs, ys)
ax.vlines(mean_black, 0, height_black, color=purple, ls=':')

mean_white = df_viz.groupby('race').age.median()[1]
height_white = np.interp(mean_white, xs, ys)
ax.vlines(mean_white, 0, height_white, color=green, ls=':')

plt.title('Distribution of Age by Race')
```

The purple curve in the plot above shows the distribution of the ages of Black defendants, and the green curve shows the distribution of the ages of white defendants. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line represents the median age of Black defendants (29 years) and the green dotted line represents the median age of white defendants (35 years). For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.

@@

When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants and Black defendants are more likely to recidivate than not recidivate. Again, this trend points to disproportionate rates of criminalization of Black people. This increased likelihood of recidivating may be attributed to the systemic racism in policing (e.g., predominantly Black neighborhoods tend to be overpoliced in comparison to predominantly white neighborhoods).

@@

@@

plt.title('Prior Charges Count by Race')
plt.show()
```

@@

plt.title('Prior Charges Count by Sex')
plt.show()
```

If a defendant has a higher number of prior offenses, the individual is more likely to be rated with a higher risk of recidivism **do we know this for sure?**. In the former plot, we can see that Black defendants have a higher number of prior charges than do white defendants. This again likely reflects the over-policing of Black people and communities. In the latter plot, we can see that male defendants have a higher number of prior charges than do female defendants. This is worth noting because the COMPAS risk assessment algorithm has been found to express a discrepancy in prediction between sexes **who found this?**. When the recidivsim predictions are compared with the true recividism outcomes, female defendants have a notable false positive rate that predicts their recidivism risk as higher than their true outcomes**where does this come from?**.

@@

$$\frac{TP}{TP + FP}$$

We can see that precision is very similar across groups, all approximately 0.6. This means that the model's predictions of a defendant, either Black or white, who recidivates are accurate about 60% of the time.

@@

$$TPR_{D = Unprivileged} - TPR_{D = Privileged}$$
```{python group fairnesss equal opportunity difference}
eq_opp_diff = equal_opportunity_difference(y_test, y_pred, prot_attr='race', pos_label = 0)

@@

This metric, an expansion of equal opportunity difference, returns the average difference in false positive rate and true positive rate for the privileged and unprivileged groups. A value of 0 is considered fair, and a value below 0 implies benefit for the privileged group. Equality of odds is achieved in the case of recidivism when the proportion of people who were predicted to recidivate and did recidivate is equal (true positive rate) for both black and white defendants AND the proportion of people who were predicted to recidivate and did not recidivate (false positive rate) is equal for both black and white defendants.

$$\frac{1}{2}\left[(FPR_{D = Unprivileged} - FPR_{D = Privileged}) + \underbrace{(TPR_{D = Unprivileged} - TPR_{D = Privileged})}_\textrm{Equal Opportunity Difference}\right]$$

@@

@@

## Reweighting

@@

The highest percentage in this confusion matrix is for the true negative rate, defendants who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model. This means that the re-weighing model marginally improved the baseline model accuracy in predicting people who did not recidivate. However, the true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model. This means that the re-weighing model slightly lowered the accuracy of the baseline model in predicting people who did recidivate. As a result, the overall reweighting model accuracy remains about the same as the baseline model accuracy. The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively. The re-weighing model increased the number of false negatives by 1.14% and decreased the number of false positives by 0.83%. We are most concerned about the false positives, which indicate defendants who are predicted to recidivate and do not actually recidivate. While there is a slight improvement in the percent of false positives, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.

@@

Like we saw in the metrics plot for the baseline model, the model accuracy for defendants of all races hovers around 61%, being slightly lower for Black defendants and slightly higher for white defendants. However, we can note that the gap between accuracy for Black and white defendants is smaller that it was with the baseline model (about a 3% gap rather than 4%).

We can see clear effects of the reweighting approach in the recall scores, which are now all very similar around 0.5. This is a marked improvement from the baseline model, where the model only correctly identified white defendants who recidivated 20% of the time. The reweighting model now correctly identifies defendants of any race who recidivate about half of the time. Notably, the reweighting also lowered the recall score for Black defendants.

Whereas with the baseline model all three precision scores were relatively similar, with the reweighting model we see the precision score for Black defendants is the highest. The models predictions for Black defendants who recidivate are correct about 66% of the time. In contrast, these predictions are correct for white defendents only 50% of the time, and correct for all defendants 60% of the time. The increase in precision for Black defendants shows the effects of the bias mitigation, as fewer Black people are being incorrectly predicted to recidivate.

@@

```{python post-processing equal odds needs fix} 
#ep = EqOddsPostprocessing(unprivileged_groups=0, privileged_groups= 1, seed =1234567)
#eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
#eo.fit(train_BLD, test_BLD)
#y_pred_EO = eo.predict(test_BLD)
#y_proba_EO = eo.predict_(test_BLD)
```
